---
title: <center>Salaries - Case Study</center>
author: <center>Shajakhan Yacob Ali</center>
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

#### $\color{Red}{\text{Loading required packages}}$
```{r "Loading Packages", message= FALSE}
library(car)
library(tidyverse)
library(ggrepel)
library(broom)
library(caret)
library(rpart)
library(rpart.plot)
library(pROC)
```
&nbsp;

#### $\color{red}{\text{Exploratory Data Analysis}}$
```{r "str & summary"}
Salaries <- as.tbl(Salaries)
str(Salaries)
summary(Salaries)
```
&nbsp;

#### $\color{green}{\text{Checking for Missing Values}}$
```{r "Missing Values"}
map_dbl(Salaries, function(x) sum(is.na(x)))
```
&nbsp;

#### $\color{green}{\text{Salary Distribution}}$
```{r "Salary Distribution"}
Salaries %>%
  ggplot(aes(salary)) +
  geom_histogram(binwidth = 12500, fill = "grey") +
  geom_vline(aes(xintercept = mean(Salaries$salary), col = "blue"), linetype = 2) +
  geom_vline(aes(xintercept = median(Salaries$salary), col = "green"), linetype = 2) +
  geom_vline(aes(xintercept = quantile(Salaries$salary, 0.25), col = "red"), linetype = 2) +
  geom_vline(aes(xintercept = quantile(Salaries$salary, 0.75), col = "red"), linetype = 2) +
  scale_color_manual("Line Color", values = c("blue" = "blue", "green" = "green", "red" = "red"),
                     labels = c("Mean", "Median", "Lower & Upper Quartiles")) +
  scale_x_continuous(breaks = seq(50000, 250000, 10000)) +
  theme(legend.position = c(0.85, 0.85), 
        legend.text=element_text(size=8),
        axis.text.x = element_text(angle = 90))

Salaries %>%
  arrange(desc(salary))
```
&nbsp;

#### $\color{green}{\text{Salary by different Categorical Variables}}$
##### $\color{blue}{\text{Salary by rank}}$
```{r "Salary by rank"}
Salaries %>%
  ggplot(aes(rank, salary)) +
  geom_boxplot()
```
&nbsp;

##### $\color{blue}{\text{Salary by discipline}}$
```{r "Salary by discipline"}
Salaries %>%
  ggplot(aes(discipline, salary)) +
  geom_boxplot()
```
&nbsp;

##### $\color{blue}{\text{Salary by sex}}$
```{r "Salary by sex"}
Salaries %>%
  ggplot(aes(sex, salary)) +
  geom_boxplot()
```
&nbsp;

##### $\color{blue}{\text{Salary by rank, sex and discipline}}$
```{r "Salary by rank, sex and discipline"}  
# Frequency Table
Salaries %>%
  group_by(sex, rank, discipline) %>%
  count()

Salaries %>%
  ggplot(aes(rank, salary)) +
  geom_boxplot() +
  facet_grid(sex ~ discipline)
```

&nbsp;

#### $\color{green}{\text{Salary vs Continuous Variables}}$
##### $\color{blue}{\text{salary vs yrs.since.phd}}$
```{r "salary vs yrs.since.phd"}  
ggplot(Salaries, aes(yrs.since.phd, salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = 4, y = 220000, parse = TRUE, col = "purple", 
           label = str_c("~r ==~", round(cor(Salaries$yrs.since.phd, Salaries$salary), 4))) +
  annotate("text", x = 4, y = 210000, parse = TRUE, col = "purple", 
           label = str_c("~r^2==~", round(cor(Salaries$yrs.since.phd, Salaries$salary) ^ 2, 4)))
```
&nbsp;

##### $\color{blue}{\text{salary vs yrs.since.phd by rank & discipline}}$
```{r "salary vs yrs.since.phd by rank, discipline"}  
Salaries %>%
  group_by(discipline, rank) %>%
  summarise(cor = round(cor(yrs.since.phd, salary), 4)) %>%
  mutate(cor = str_c("r = ", cor)) %>%
  as.data.frame() -> cor_yrs.since.phd_salary

ggplot(Salaries, aes(yrs.since.phd, salary, col = rank, shape = discipline)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(rank ~ discipline) +
  geom_text(data = cor_yrs.since.phd_salary, aes(x = 0, y = 220000, label = cor), 
            hjust = "inward", size = 3, col = "purple", inherit.aes = FALSE) +
  theme(legend.position = "none")
```
&nbsp;

##### $\color{blue}{\text{salary vs yrs.service}}$
```{r "salary vs yrs.service"}  
ggplot(Salaries, aes(yrs.service, salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = 4, y = 220000, parse = TRUE, col = "purple", 
           label = str_c("~r ==~", round(cor(Salaries$yrs.service, Salaries$salary), 4))) +
  annotate("text", x = 4, y = 210000, parse = TRUE, col = "purple", 
           label = str_c("~r^2==~", round(cor(Salaries$yrs.service, Salaries$salary) ^ 2, 4)))
```
&nbsp;

##### $\color{blue}{\text{Salary vs yrs.service by rank & discipline}}$
```{r "Salary vs yrs.service by rank & discipline"}  
Salaries %>%
  group_by(discipline, rank) %>%
  summarise(cor = round(cor(yrs.service, salary), 4)) %>%
  mutate(cor = str_c("r = ", cor)) %>%
  as.data.frame() -> cor_yrs.service_salary

ggplot(Salaries, aes(yrs.service, salary, col = rank, shape = discipline)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(rank ~ discipline) +
  geom_text(data = cor_yrs.service_salary, aes(x = 0, y = 220000, label = cor), 
            hjust = "inward", size = 3, col = "purple", inherit.aes = FALSE) +
  theme(legend.position = "none")
```

&nbsp;

##### $\color{green}{\text{yrs.service vs yrs.since.phd}}$
```{r "yrs.service vs yrs.since.phd"}  
ggplot(Salaries, aes(yrs.since.phd, yrs.service))+
  geom_point()+
  scale_x_continuous(breaks = seq(0,60,5)) + 
  scale_y_continuous(breaks = seq(0,65,5)) +
  geom_abline(slope =1, intercept=0) +
  annotate("text", x = 5, y = 55, parse = TRUE, col = "purple", 
           label = str_c("~r ==~", round(cor(Salaries$yrs.since.phd, Salaries$yrs.service), 4)))
```

&nbsp;

#### $\color{Red}{\text{Analysis Task}}$
#####1. What percentage of records are Assistant Professors with less than 5 years of experience?
```{r "Question 1"}
mean(Salaries$rank == "AsstProf" & Salaries$yrs.service < 5)
```
$\color{green}{\text{15.87% of records are Assistant Professors with less than 5 years of experience}}$

&nbsp;

#####2. Is there a statistically significant difference between female and male salaries?
```{r "Question 2"}
wilcox.test(salary ~ sex, data = Salaries)
```
$\color{green}{\text{p-value of 0.008237 confirms that the difference in Male and Female salaries are statistically significant}}$

&nbsp;

#####3. What is the distribution of salary by rank and discipline?
```{r "Question 3"}
Salaries %>%
  group_by(discipline, rank) %>%
  mutate(Minimum = min(salary),
         Lower_Quartile = quantile(salary, 0.25),
         Median = median(salary),
         Mean = mean(salary),
         Upper_Quartile = quantile(salary, 0.75),
         Maximum = max(salary)) %>%
  ggplot(aes(salary, group = interaction(discipline, rank))) +
  geom_density() +
  geom_vline(aes(xintercept = Mean, col = "blue"), linetype = 2) +
  geom_vline(aes(xintercept = Median, col = "green"), linetype = 2) +
  geom_vline(aes(xintercept = Lower_Quartile, col = "red"), linetype = 2) +
  geom_vline(aes(xintercept = Upper_Quartile, col = "red"), linetype = 2) +
  facet_wrap(discipline ~ rank, scales = "free", labeller = label_context) +
  scale_color_manual("Line Color", values = c("blue" = "blue", "green" = "green", "red" = "red"),
                     labels = c("Mean", "Median", "Lower & Upper Quartiles")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")
```

&nbsp;

#####4. How would you recode discipline as a 0/1 binary indicator?
```{r "Question 4"}
Salaries %>%
  mutate(row = row_number(),
         discipline = str_c("discipline_",discipline),
         value = 1) %>%
  spread(discipline, value, fill = 0) %>%
  select(-row) %>%
  str()
```
&nbsp;

#### $\color{red}{\text{Build a predictive model using salary as the response}}$
##### $\color{blue}{\text{Linear Model 1}}$
```{r "Linear Model 1"}
linear_mod_1 <- lm(salary ~ . -sex, data = Salaries)
summary(linear_mod_1)
```
**Note:** I have excluded sex because it is not statistically significant when added with other variables

&nbsp;

##### $\color{green}{\text{Model Performance}}$
$\color{purple}{\text{Adjusted R-squared = 0.4455}}$  
yrs.since.phd, yrs.service, rank and discipline explain 44.55% of change in salary.  
&nbsp;

$\color{purple}{\text{Residual standard error = 22550}}$  
The typical difference between the observed salary and the salary predicted by the model is about $22550.  
&nbsp;


#### $\color{blue}{\text{Linear Model 2}}$
From my exploratory data analysis, I know that yrs.since.phd and yrs.service have week co-relation with salary especially when grouped by discipline and rank. 
```{r "Linear Model 2"}
linear_mod_2 <- lm(salary ~ discipline + rank, data = Salaries)
summary(linear_mod_2)
```

&nbsp;

##### $\color{green}{\text{Model Performance}}$
$\color{purple}{\text{Adjusted R-squared = 0.4407}}$  
rank and discipline explain 44.07% of change in salary.  
&nbsp;

$\color{purple}{\text{Residual standard error = 22650}}$  
The typical difference between the observed salary and the salary predicted by the model is about $22650.  

**Note: This model's Adjusted R-squared and Residual standard error are very close to linear_mod_1. Thus, we can conclude that yrs.since.phd & yrs.service have very little to no impact on salary **

&nbsp;

##### $\color{green}{\text{Predicted Salary}}$
$\color{purple}{\text{salary = 13762 * is_associate_prof + 47844 * is_prof + 13761 * is_discipline_B + 71944}}$  

* Salary for associate professor is $13762 more than assistant professor considering they have same discipline
* Salary for professor is $47844 more than assistant professor considering they have same discipline
* Salary for someone in discipline B is $13761 more than someone in discipline A considering they have same rank 
```{r "Predicted Salary"}  
linear_mod_2_aug <- augment(linear_mod_2) %>%
  mutate(row_number = row_number())

linear_mod_2_aug %>%
  transmute(discipline, rank, Predicted_Salary = .fitted) %>%
  distinct() %>%
  spread(rank, Predicted_Salary)
```

&nbsp;

##### $\color{green}{\text{Influential Points}}$  
```{r "Influential Points Graph"}
ggplot(linear_mod_2_aug, aes(row_number, .cooksd)) +
  geom_point(alpha = 0.5) +
  geom_hline(aes(yintercept = mean(.cooksd), col = "green"), linetype = 2) +
  geom_hline(aes(yintercept = (4 * mean(.cooksd)), col = "red"), linetype = 2) +
  geom_text_repel(aes(label = if_else(.cooksd > (4 * mean(.cooksd)), as.character(row_number), "")), 
                  col = "red", size = 3) +
  scale_color_manual("Line Color", values = c("green" = "green", "red" = "red"),
                     labels=c("Mean", "4 * Mean"))
```

```{r "Influential Points"}
linear_mod_2_aug %>%
  filter(.cooksd > 4 * mean(.cooksd)) %>% 
  arrange(desc(.cooksd)) %>%
  select(-c(.hat, .sigma, .se.fit, .std.resid)) %>%
  as.data.frame()
```

**Note:** This data should be analysed further (preferably with SME) and appropriate action should be taken before finalizing the model

&nbsp;

##### $\color{green}{\text{residual vs fitted}}$  
```{r "residual vs fitted"}
ggplot(linear_mod_2_aug, aes(.fitted, .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(aes(yintercept = 0))
```

**Note:** The resid vs fitted values plot shows that the model is relatively accurate when it predicts low values of salary, and highly inaccurate when it predicts high values of salary.

&nbsp;

#### $\color{red}{\text{Build a model to predict binary response}}$
##### $\color{green}{\text{Creating 0/1 binary indicator using salary}}$
```{r}
(Salaries_binary <- Salaries %>%
  mutate(salary_binary = if_else(salary < median(salary), 0 , 1)) %>%
  as.tbl())
```
&nbsp;

##### $\color{green}{\text{Splitting the data into 70% for training and 30% for testing the model}}$
```{r}
set.seed(1)
sample_rows <- sample(nrow(Salaries_binary), nrow(Salaries_binary) * 0.7)

Salaries_binary_train <- Salaries_binary[sample_rows,]
Salaries_binary_test <- Salaries_binary[-sample_rows,]
```
&nbsp;

#### $\color{blue}{\text{1. Logistic Regression:}}$
```{r}
logistic_model <- glm(salary_binary ~ discipline + rank , 
                      data = Salaries_binary_train, family = "binomial")
```
&nbsp;

##### $\color{green}{\text{Predicting salary_binary using test data}}$
```{r}
Salaries_logistic_test_out <- Salaries_binary_test %>%
  mutate(outcome_prob = predict(logistic_model, Salaries_binary_test, type = "response"),
         outcome_binary = if_else(outcome_prob > 0.5, 1, 0))
```
&nbsp;

$\color{purple}{\text{Probability Prediction}}$
```{r}
Salaries_logistic_test_out %>%
  distinct(discipline, rank, outcome_prob) %>%
  spread(rank, outcome_prob)
```
&nbsp;

$\color{purple}{\text{Binary Prediction (Threshold: >0.5)}}$
```{r}
Salaries_logistic_test_out %>%
  distinct(discipline, rank, outcome_binary) %>%
  spread(rank, outcome_binary)
```
&nbsp;

##### $\color{green}{\text{Model Performance}}$
```{r}
confusionMatrix(as.factor(Salaries_logistic_test_out$outcome_binary), 
                as.factor(Salaries_logistic_test_out$salary_binary),
                positive = "1")
```
&nbsp;

##### $\color{green}{\text{ROC curve}}$
```{r}
logistic_model_roc <- roc(Salaries_logistic_test_out$salary_binary, Salaries_logistic_test_out$outcome_binary)
plot(logistic_model_roc, col = "blue", print.auc = TRUE)
```
&nbsp;

#### $\color{blue}{\text{2. Decision Tree:}}$
```{r}
decision_tree_model <- rpart(salary_binary ~ discipline + rank, data = Salaries_binary_train, 
                                      method = "class", control = rpart.control(cp = 0))

rpart.plot(decision_tree_model, type = 4)
```

##### $\color{green}{\text{Predicting salary_binary using test data}}$
```{r}
Salaries_decisionTree_test_out <- Salaries_binary_test %>%
  mutate(outcome_binary = predict(decision_tree_model, Salaries_binary_test, type = "class"),
         outcome_binary = as.numeric(as.character(outcome_binary)))

Salaries_decisionTree_test_out %>%
  distinct(discipline, rank, outcome_binary) %>%
  spread(rank, outcome_binary)
```
&nbsp;

##### $\color{green}{\text{Model Performance}}$
```{r}
confusionMatrix(as.factor(Salaries_decisionTree_test_out$outcome_binary),
                as.factor(Salaries_decisionTree_test_out$salary_binary),
                positive = "1")
```
&nbsp;

##### $\color{green}{\text{ROC curve}}$
```{r}
decisionTree_roc <- roc(Salaries_decisionTree_test_out$salary_binary, Salaries_decisionTree_test_out$outcome_binary)
plot(decisionTree_roc, col = "blue", print.auc = TRUE)
```
&nbsp;

#### $\color{red}{\text{Dataset Enhancement}}$
**1. State at least three research questions you would like to address and describe your thought process behind how you formulated these research questions**

  1. Is the salary difference across different ethnicity and race decreasing over the last 10 years?  
     To find if the efforts to reduce ethnic and racial inequality is working or not for this profession.
     
  2. Is Number of international papers authored/co-authored a good predictor of salary?  
     In addition to thier usual work, Professors continue to author or co-author papers/journals. Is this one of the variables used to measure their performance and does this get them a pay raise? 
     
  3. How does salary vary with the quality of benefits offered?  
     To find if there is a decreasing trend between quality of benefits offered and salary.   

**2. Prepare a list of 5-7 additional attributes you would like to add to the data set. Prepare a brief explanation for each attribute**  

  1. ethnicity (fct) : Ethnicity of the professor
  2. race (fct) : Race of the professor
  3. year (num) : Year from 2009 - 2017
  4. salary_year (num) : Salary corresponding to the year variable
  5. international_papers_year (num) : Total number of international papers authored/co-authored corresponding to the year variable
  6. vacation_days (num) : Total number of vacation days per year
  7. health_insurance_quality (fct) : 1 - no insurance; 2 - poor; 3 - moderate; 4 - good; 5 - excellent
  8. retirement_benefit_quality (fct) : 1 - no benefit; 2 - poor; 3 - moderate; 4 - good; 5 - excellent

**3. Estimate and justify the appropriate sample size (and sampling technique, if desired) that would be required to address the research questions you defined**
  
  **Sampling size**  
  I have no experience on this subject and my answer is based on my research on this topic.  
  
  Sample size is usually determined by the following factors: 
  
  1. Population size  
  2. Confidence level  
  3. Margin of error  
    
  There are Several available tools and methods to calculate the sample size using the above mentioned factors. One such method that i found during my research is Cochran Formula for Sample Size Calculation.
  
  **Sampling Technique**  
  
  1. Is the salary difference across different ethnicity and race decreasing over the last 10 years?es?  
     Sampling Technique: Stratified Sampling (Groups: Ethnicity & Race)  
     
  2. Is Number of international papers authored/co-authored a good predictor of salary?  
     Sampling Technique: Random Sampling  
     
  3. How does salary vary with the quality of benefits offered?  
     Sampling Technique: Random Sampling 